A Splitting Solution of the Linearity Challenge

### 2025-12-11 paper/Process.lagda.tex
revised description of processes

\subsection{Process Representation}
\label{sec:process-agda}

@@

The constructor \AgdaInductiveConstructor{fail} requires the context $\Context$
to satisfy the predicate
$\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊤}~\AgdaDatatype{∗}~\AgdaFunction{U}$,
meaning that $\Context$ must contain an entry \AgdaInductiveConstructor{⊤}. That
form of $\Context$ matches the one in the conclusion of \FailRule.

@@

Crucially, the type of the continuation channels are \emph{prepended} to the
context that indexes the continuation processes using the
\AgdaInductiveConstructor{\_∷\_} list constructor. This is consistent with the
fact that the continuation channels have been freshly introduced, and the
channel on which the process operates has been consumed.
%
There are just a few deviations from this pattern, which we discuss next.

The constructor \AgdaInductiveConstructor{fork}, which represents a process
$\Fork\x\y\z{P}{Q}$, has an additional splitting argument
$\Splitting\ContextD{\Context_1}{\Context_2}$ which distributes the resources in
$\ContextD$ among the two continuation processes $P$ and $Q$.

The constructor \AgdaInductiveConstructor{all}, which represents a process
$\Forall\x\X\y.P$, models the continuation $P$ using higher-order abstract
syntax (HOAS~\cite{}): the continuation $P$ is produced by a function applied to
a \AgdaInductiveConstructor{Type} $X$. Notice how the constructor enforces the
side condition of \ForallRule \emph{by definition}:
%
$A$ has type
$\AgdaDatatype{PreType}~(\AgdaInductiveConstructor{suc}~\AgdaInductiveConstructor{zero})$,
that is a pre-type with \emph{at most one} free type variable (that is $X$),
whereas $\ContextD$ is a list of $\AgdaDatatype{Type} =
\AgdaDatatype{PreType}~\AgdaInductiveConstructor{zero}$, that is a list of
pre-types without free type variables. Therefore, $X$ cannot occur in
$\ContextD$ as required by \ForallRule.

Finally, the constructor \AgdaInductiveConstructor{server}, which represents a
process $\Server\x\y.P$, has an additional argument of type
$\AgdaDatatype{Un}~\ContextD$ enforcing the property that $\ContextD$ is
unrestricted.

The two remaining constructors, \AgdaInductiveConstructor{link} and
\AgdaInductiveConstructor{cut}, deal with links and cuts in the expected way. A
link is well typed in a context
$\Splitting\Context{[A]}{[\AgdaFunction{dual}~A]}$. We use a splitting instead
of requiring $\Context$ to be $[A,\AgdaFunction{dual}~A]$ so that it becomes
straightforward to define \SLink using \AgdaFunction{+-comm}.
%
The constructor \AgdaInductiveConstructor{cut} models a process $\Cut\x{P}{Q}$
and uses splitting to distribute the resources in $\Context$ among $P$ and $Q$.

Permutations preserve process typeability. In a sense we can read this property
as the fact that we can rename channels at will.

>>>>

\subsection{Channel and Process Representation}
\label{sec:process-agda}

@@

For example, the constructor \AgdaInductiveConstructor{fail}, which corresponds
to a process of the form $\Fail\x$, requires the context $\Context$ to satisfy
the predicate
$\AgdaDatatype{Ch}~\AgdaInductiveConstructor{⊤}~\AgdaDatatype{∗}~\AgdaFunction{U}$,
meaning that $\Context$ must contain an entry \AgdaInductiveConstructor{⊤}. That
form of $\Context$ matches the one in the conclusion of \FailRule.

@@

Let us move on to the processes that produce a continuation channel. For
example, the constructor \AgdaInductiveConstructor{case} corresponds to a
process of the form $\Case\x\y{P}{Q}$, where both $P$ and $Q$ may use the
continuation channel $y$. In this case $\Context$ must satisfy the predicate
\[
  \AgdaDatatype{Ch}~(A~\AgdaInductiveConstructor{\&}~B)~\AgdaDatatype{∗}~((A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}~\AgdaFunction{∩}~(B~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc})
\]
which looks intimidating at first but makes perfect sense once we recall
\CaseRule. Remember that we are trying to establish whether $\Case\x\y{P}{Q}$ is
well typed in $\Context$.
%
The predicate $\AgdaDatatype{Ch}~(A~\AgdaInductiveConstructor{\&}~B)$ expresses
the requirement that the type of $x$ must be of the form $A \With B$ and should
be found in $\Context$. In other words, it must be the case that
$\Splitting\Context{[A \With B]}\ContextD$ for some $\ContextD$. Now $P$ and $Q$
must be well typed in the context $\ContextD$ augmented with the association $y
: A$ and $y : B$, respectively, whence the use of \AgdaFunction{∩} to verify a
(non-separating) conjunction of the predicates
$(A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}$ and
$(B~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}$. The
two new contexts are obtained by \emph{adding} either $A$ or $B$ to $\ContextD$,
which we perform using \AgdaFunction{⊢}.
%
Crucially, the types $A$ and $B$ are \emph{prepended} to $\ContextD$, which is
consistent with the fact that $y$ has been freshly introduced and $x$ has been
consumed. Notice how easy it is to \emph{prepend} either $A$ or $B$ to
$\ContextD$ instead of \emph{changing} the type of $x$ in $\Context$ from $A
\With B$ to either $A$ or $B$ while preserving the position of the type, as we
would have to do in a ``true'' session type system without explicit continuation
channels.

The interpretation of the remaining constructors is analogous, so we only
comment \AgdaInductiveConstructor{all}, which represents a process
$\Forall\x\X\y.P$. This constructor models the continuation $P$ using
higher-order abstract syntax (HOAS~\cite{}): witnessing that a context
$\Context$ satisfies the predicate
\[
  \AgdaFunction{⋂[}~X~\AgdaFunction{∶}~\AgdaDatatype{Type}~\AgdaFunction]~((\AgdaFunction{subst~[}~X~\AgdaFunction{/\_]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
means providing a function that, for every type $X$, produces a witness for the
predicate
\[
  ((\AgdaFunction{subst~[}~X~\AgdaFunction{/\_]}~A~\AgdaInductiveConstructor{∷\_})~\AgdaFunction{⊢}~\AgdaDatatype{Proc}
\]
applied to $\Context$. Note that the side condition $X \not\in \fv\Context$ of
\ForallRule is enforced \emph{by definition}:
%
$A$ has type $\AgdaDatatype{PreType}~\AgdaInductiveConstructor{1}$, that is a
pre-type with \emph{at most one} free type variable ($X$), whereas $\Context$ is
a typing context, that is a list of $\AgdaDatatype{Type} =
\AgdaDatatype{PreType}~\AgdaInductiveConstructor{0}$ without free type
variables. Therefore, $X$ cannot occur in $\Context$.

We conclude this module proving that permutations preserve process typeability.
Since list permutations basically correspond to channel renaming, we can read
this property as the fact that typeability is preserved by (injective) name
substitutions.\Luca{Abbiamo detto da qualche parte dove emergono gli indici di
de Bruijn e di fare attenzione al fatto che lo splitting li altera? Serve questa
digressione?}\Luca{Da qualche parte mettere un remark per far notare che l'uso
della separating conjunction introduce rumore, ma avere i costruttori
currificati ne complica un po' l'interpretazione dei tipi. Forse è meglio fare
questo rilievo nella prossima sezione quando è evidente il pattern matching che
serve per esprimere le regole.}
